Future challenges for Research Software Engineering

### 2025-08-01 future_challenges_for_RSEng.tex
Break long lines, sort BibTeX records

% Required packages
% =================
% Your \usepackage commands go here.
\usepackage{fancybox,framed}
\usepackage{footnote}
\makesavenoteenv[FramedParagraphWithFootnotes]{framed} % for author notes
\usepackage{color}
\usepackage{orcidlink}
\usepackage[dvipsnames]{xcolor}
\usepackage[most]{tcolorbox}

@@

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are common in industry~\cite{gartner}, so far there is a projection lacking for the RSEng space.
We present here the results of a workshop discussion at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how Research Software Engineering should, or could, develop.
It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software~\cite{Hasselbring2024}.
It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.

In order to structure the discussion, we decided to consider
the evolution of the term RSE and its definition,
the growth and development of the RSE community,
the impact of automation and generative artificial intelligence in RSE work,
the impact of complexity and usability in research software,
ethical and social considerations,
the expected transformations in academic institutions,
and the potential impact of alternative career paths for RSEs and new groups of RTPs (\emph{Research Technical Professional}s).
%
We conclude the paper with a summary and with our wishes and recommendations.

@@

\begin{story}{The Definition of RSEs}{defrse}
Kim and Kay chuckled when they looked at their original job descriptions from 2020: "build tools for scientists, manage data, write reusable code."
Now, their roles are unrecognizable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who transformed the chaos back into meaningful order.
They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.
Their job titles have changed, but their role as bridges between logic and life remains the same.
In a future of shifting code and constant reinvention, it is their understanding of people that never went out of date.

@@

\subsection{Development of the RSE community}
\begin{whatis}{}{development-rse-community}
National RSE communities are increasingly collaborating with local partners, fostering stronger networks across countries. Regular international conferences further amplify these connections, promoting the global exchange of ideas. The introduction of RSE Master's programmes, coupled with a shift in culture towards valuing Research Software, has significantly elevated its profile. This growing recognition has led to a surge in membership of national societies, reflecting the increasing importance of RSEng in the research landscape.
\end{whatis}
In recent years, we have seen the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation%
of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSE positions to become more easily accessible, possibly through centrally-funded long-term positions at research institutions.
The community of RSEs will get a significant boost in their acceptance if RSE Master's programmes
become available at various universities.
With increased numbers, the visibility of the RSE community to industry will be enhanced.
\begin{story}{Development of RSE community}{RSEcommunity}
Kay stands proudly at the podium of the first International Research Software Engineering Conference held in the Southern Hemisphere, a historic moment for both her career and the global RSE community. Having moved to South Africa, she speaks passionately about the unique challenges researchers in the Global South face: limited access to high-performance computing, unreliable funding, and the persistent digital divide. Her keynote highlights how local innovation thrives despite constraints, often out of necessity rather than choice. The audience, a diverse mix of global experts, listens intently as she calls for equitable collaborations, not just inclusion as an afterthought.
\end{story}

@@

Another risk of the widespread use of AI tools is that community resources such as \eg Stack Overflow and Wikipedia become poorer because there are fewer incentives to maintain them.

@@

% society implications
The work being undertaken by RSEs may well have direct, or indirect,
ethical implications in areas such as social engineering~\cite{s2erc,siadati2024},
climate~\cite{Lannelongue2023} and more.
Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
As engineers, RSEs take the responsibility for the software they develop.
RSEs need to be aware of these issues and develop their own ``red lines''.
With the advantages of \autoref{sec:ai}, ever-more AI created artifacts will be included in research.
As a result of this, we expect a discussion within societies about their acceptable use.

@@

\subsection{Impact of Alternative Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{}{career-paths}
We would like to see the emergence of more dedicated staff, i.e. other (digital) RTPs next to the RSE.
Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg in think tanks, NGOs,
or organisations that do knowledge transfer).
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units~\cite{Kempf2025-draft}.
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding can be more difficult to acquire in lower tiers of a research organisation (i.e., at departmental or individual research group level).
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, and Göttingen, as well as several universities in the UK where the concept of dedicated RSE groups began to emerge around 2013).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups prove themselves to be increasingly valuable and, hence, sustainable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.

>>>>

% Required packages
% =================
% Your \usepackage commands go here.
\usepackage[T1]{fontenc}
\usepackage{fancybox,framed}
\usepackage{footnote}
\makesavenoteenv[FramedParagraphWithFootnotes]{framed} % for author notes
\usepackage{color}
\usepackage{orcidlink}
\usepackage[dvipsnames]{xcolor}
\usepackage[most]{tcolorbox}

@@

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are common in industry~\cite{Gartner2025},
so far there is a projection lacking for the RSEng space.
We present here the results of a workshop discussion at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how Research Software Engineering should, or could, develop.

It is already difficult to structure the spectrum of RSE tasks and responsibilities
today given the spectrum of research software~\cite{Hasselbring2024}.
It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
In order to structure the discussion, we decided to consider
the evolution of the term RSE and its definition,
the growth and development of the RSE community,
the impact of automation and generative artificial intelligence in RSE work,
the impact of complexity and usability in research software,
ethical and social considerations,
the expected transformations in academic institutions,
and the potential impact of alternative career paths for RSEs
and new groups of RTPs (\emph{Research Technical Professional}s).
We conclude the paper with a summary and with our wishes and recommendations.

@@

\begin{story}{The Definition of RSEs}{defrse}
Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
Now, their roles are unrecognizable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who transformed the chaos back into meaningful order.
They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.
Their job titles have changed, but their role as bridges between logic and life remains the same.
In a future of shifting code and constant reinvention, it is their understanding of people that never went out of date.

@@

\subsection{Development of the RSE community}
\begin{whatis}{}{development-rse-community}
National RSE communities are increasingly collaborating with local partners,
fostering stronger networks across countries.
Regular international conferences further amplify these connections,
promoting the global exchange of ideas. The introduction of RSE Master's programmes,
coupled with a shift in culture towards valuing Research Software,
has significantly elevated its profile.
This growing recognition has led to a surge in membership of national societies,
reflecting the increasing importance of RSEng in the research landscape.
\end{whatis}
In recent years, we have seen the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSE positions to become more easily accessible,
possibly through centrally-funded long-term positions at research institutions.
The community of RSEs will get a significant boost in their acceptance
if RSE Master's programmes become available at various universities.
With increased numbers, the visibility of the RSE community to industry will be enhanced.
\begin{story}{Development of RSE community}{RSEcommunity}
Kay stands proudly at the podium of the first International Research
Software Engineering Conference held in the Southern Hemisphere,
a historic moment for both her career and the global RSE community.
Having moved to South Africa, she speaks passionately about the unique
challenges researchers in the Global South face: limited access
to high-performance computing, unreliable funding,
and the persistent digital divide.
Her keynote highlights how local innovation thrives despite constraints,
often out of necessity rather than choice.
The audience, a diverse mix of global experts, listens intently as she calls
for equitable collaborations, not just inclusion as an afterthought.
\end{story}

@@

Another risk of the widespread use of AI tools is that community resources
such as \eg\ Stack Overflow and Wikipedia become poorer because there are fewer incentives to maintain them.

@@

% society implications
The work being undertaken by RSEs may well have direct, or indirect,
ethical implications in areas such as social engineering~\cite{s2erc,Siadati2024},
climate~\cite{Lannelongue2023} and more.
Another example might be that they are developing code that can also be used
for military purposes, such as numerical solvers or computer vision systems.
As engineers, RSEs take the responsibility for the software they develop.
RSEs need to be aware of these issues and develop their own ``red lines''.
With the advantages of \autoref{sec:ai}, ever-more AI created artifacts will be included in research.
As a result of this, we expect a discussion within societies about their acceptable use.

@@

\subsection{Impact of Alternative Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{}{career-paths}
We would like to see the emergence of more dedicated staff, i.e. other (digital) RTPs next to the RSE.
Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg\ in think tanks, NGOs,
or organisations that do knowledge transfer).
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units~\cite{Kempf2025-draft}.
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding can be more difficult to acquire in lower tiers
of a research organisation (i.e., at departmental or individual research group level).
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, and Göttingen, as well as several universities
in the UK where the concept of dedicated RSE groups began to emerge around 2013).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups prove themselves to be increasingly valuable and, hence, sustainable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.
